Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2014-01-17 acm/ieee_paper.tex
line number

\def\red#1{\textbf{\textcolor{red}{#1}}}
\def\blue#1{\textbf{\textcolor{blue}{#1}}}
\def\green#1{\textbf{\textcolor{green}{#1}}}
\def\qes#1{{\blue{*** For Erik: #1 ***}}}
\def\es#1{{\blue{*** For Erik: #1 ***}}}
\def\ee#1{{\green{#1}}}
\def\ge#1{{\blue{#1}}}
%\def\NOTE#1{{\red{NOTE: #1}}}
\def\NOTE#1{{}}
\def\ttt#1{{\tt #1}}
\def\bold#1{{\bf #1}}

@@

\ee{The core of the technique is to load the maximum amount of data
  into large register and format the data for processing within these
  registers. That way the amount of interaction with the memory is
  minimized. The Intel MIC is more for these operations than classical
  CPUs because it features larger vector registers and more elaborate
  vector reorganization instructions than namely swizzling,
  permutations and broadcasts. (The AVX instruction sets on classical
  CPUs is more primitive in that aspect.)} A vector of 512 bits is
composed of four lanes (sometimes called channels) of 128 bits, which
are made of four segments of 32 bits. A swizzling operation allows the
reorder of the elements within each lane (see how {\tt mat} is
extracted from {\tt mat\_ent} in Figure~\ref{code:mat_mul}). On the
other hand, a permutation reorders entire lanes, without changing
their contents (see for instance how {\tt vect} is generated from {\tt
  all\_vect} in Figure~\ref{code:mat_mul}).  Both permutation and
swizzling support common reordering and broadcast. In the code, the
``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert a
vector to and from integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permutation is its own
instruction.)

@@

Loading the entries from the {\tt x} vector is  more
involved. First, the column pointers of four different nonzeros are
loaded into a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to {\tt unpack} for the
LSB and MSB. Both are mandatory even if one is known to be
nilpotent according the documentation of the hardware.} (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
are replicated so that each value fills a single vector lane 
using a swizzle operation. Then each value is multiplied by four
(because $n_v=4$) 
and the offset vector $(3,2,1,0)$ is added to each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation. A {\tt gather} operation is then
performed to bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a register where each lane is the
four vector entries for a nonzero element. Using a broadcast
permutation, one can replicate an entry of the four vectors across
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul}.)

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the "No Read No Global
Order" hint to optimize write traffic on the memory bus.

>>>>

\def\red#1{\textbf{\textcolor{red}{#1}}}
\def\blue#1{\textbf{\textcolor{blue}{#1}}}
\def\green#1{\textbf{\textcolor{green}{#1}}}
\def\qes#1{{\blue{*** For Erik: #1 ***}}}
\def\es#1{{\blue{*** For Erik: #1 ***}}}
\def\ee#1{{\green{#1}}}
\def\ge#1{{\blue{#1}}}
\def\NOTE#1{{\red{NOTE: #1}}}
%\def\NOTE#1{{}}
\def\ttt#1{{\tt #1}}
\def\bold#1{{\bf #1}}

@@

\ee{The core of the technique is to load the maximum amount of data
  into large register and format the data for processing within these
  registers. That way the amount of interaction with the memory is
  minimized. The Intel MIC is more for these operations than classical
  CPUs because it features larger vector registers and more elaborate
  vector reorganization instructions than namely swizzling,
  permutations and broadcasts. (The AVX instruction sets on classical
  CPUs is more primitive in that aspect.)} A vector of 512 bits is
composed of four lanes (sometimes called channels) of 128 bits, which
are made of four segments of 32 bits. A swizzling operation allows the
reorder of the elements within each lane (see how {\tt mat} is
extracted from {\tt mat\_ent} in Figure~\ref{code:mat_mul} VIII to X). On the
other hand, a permutation reorders entire lanes, without changing
their contents (see for instance how {\tt vect} is generated from {\tt
  all\_vect} in Figure~\ref{code:mat_mul} VII to IX).  Both permutation and
swizzling support common reordering and broadcast. In the code, the
``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert a
vector to and from integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permutation is its own
instruction.)

@@

Loading the entries from the {\tt x} vector is  more
involved. First, the column pointers of four different nonzeros are
loaded into a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation in line IV\footnote{Notice that there are two calls to {\tt unpack} for the
LSB and MSB. Both are mandatory even if one is known to be
nilpotent according the documentation of the hardware.} (The column
pointers are named $j,k,l,m$ in Figure~\ref{code:mat_mul}.) They
are replicated so that each value fills a single vector lane 
using a swizzle operation (in line V). Then each value is multiplied by four
(because $n_v=4$) 
and the offset vector $(3,2,1,0)$ is added to each lane to obtain the
correct index of the elements of the {\tt x} vector within each lane
using a Fused Multiply-Add operation in line VI. A {\tt gather} operation is then
performed in line VII to bring the 16 entries of the {\tt x} vector into the SIMD
register. At this point we have a register where each lane is the
four vector entries for a nonzero element. Using a broadcast
permutation, one can replicate an entry of the four vectors across
each lane. (The broadcast of $x_m^1,x_m^2,x_m^3,x_m^4$ is shown in
Figure~\ref{code:mat_mul} in line IX.)

Now that $A_h^1,A_h^2,A_h^3,A_h^4$ and $x_m^1,x_m^2,x_m^3,x_m^4$ are
properly laid out in the vectors, a partial value for $y_i$ can be
computed in line XI. Once all the nonzeros of the row have been processed, the
exact $y_i$ value can be sent to memory using a the "No Read No Global
Order" hint to optimize write traffic on the memory bus.
